---
title: "World Happiness in Southeast Asia"
subtitle: "An analysis of World Happiness Report data for the 11 Southeast Asian countries (R)"
format:
  html:
    toc: true
    code-fold: true
    code-summary: "Show code"
    embed-resources: true
    fig-width: 8
    fig-height: 5
execute:
  warning: false
  message: false
params:
  data_files: "DataForTable2.1.xls"
---

## Setup

Set `data_files` in the YAML header to your World Happiness Report data, or pass
it when rendering. Separate several files with commas. Either layout works:

- **Panel data** (all years): `DataForTable2.1.xls` from
  <https://worldhappiness.report/data/>, or Kaggle's `world-happiness-report.csv`.
- **Single-year files**: Kaggle's `world-happiness-report-2021.csv`, or
  `2015.csv` ... `2019.csv`. List several and the year is read from each file name.

```{r}
#| label: setup
library(dplyr)
library(tidyr)
library(readr)
library(readxl)
library(stringr)
library(purrr)
library(ggplot2)
library(knitr)

data_files <- str_trim(str_split_1(params$data_files, ","))

# Canonical name -> spellings used across WHR editions / other datasets.
sea_countries <- list(
	"Brunei" = c("brunei", "brunei darussalam"),
	"Cambodia" = c("cambodia"),
	"Indonesia" = c("indonesia"),
	"Laos" = c("laos", "lao pdr", "lao people's democratic republic"),
	"Malaysia" = c("malaysia"),
	"Myanmar" = c("myanmar", "burma"),
	"Philippines" = c("philippines", "the philippines"),
	"Singapore" = c("singapore"),
	"Thailand" = c("thailand"),
	"Timor-Leste" = c("timor-leste", "east timor", "timor leste"),
	"Vietnam" = c("vietnam", "viet nam")
)
alias_to_country <- setNames(
	rep(names(sea_countries), lengths(sea_countries)),
	unlist(sea_countries, use.names = FALSE)
)

# Standard column -> names it appears under in different WHR files.
column_aliases <- list(
	country = c("country name", "country", "country or region"),
	year = c("year"),
	score = c("life ladder", "ladder score", "life evaluation",
		"happiness score", "score"),
	gdp = c("log gdp per capita", "logged gdp per capita",
		"economy (gdp per capita)", "gdp per capita"),
	social_support = c("social support", "family"),
	life_expectancy = c("healthy life expectancy at birth",
		"healthy life expectancy", "health (life expectancy)"),
	freedom = c("freedom to make life choices", "freedom"),
	generosity = c("generosity"),
	corruption = c("perceptions of corruption",
		"trust (government corruption)")
)

factor_labels <- c(
	score = "Happiness score",
	gdp = "GDP per capita",
	social_support = "Social support",
	life_expectancy = "Healthy life expectancy",
	freedom = "Freedom of choice",
	generosity = "Generosity",
	corruption = "Perceived corruption"
)
all_factors <- names(factor_labels)[-1]

# Chart colours: one hue for magnitude, a two-pole pair for +/- correlation.
blue <- "#2a78d6"
orange <- "#eb6834"
gray <- "#b5b3ad"
text_muted <- "#52514e"

theme_set(
	theme_minimal(base_size = 11) +
	theme(
		panel.grid.minor = element_blank(),
		panel.grid.major = element_line(colour = scales::alpha(gray, 0.35)),
		axis.text = element_text(colour = text_muted),
		axis.title = element_text(colour = text_muted),
		plot.title.position = "plot"
	)
)

load_file <- function(path) {
	df <- if (str_detect(tolower(path), "\\.xlsx?$")) {
		read_excel(path)
	} else {
		read_csv(path, show_col_types = FALSE)
	}
	lookup <- setNames(names(df), str_trim(tolower(names(df))))
	out <- list()
	for (std in names(column_aliases)) {
		hit <- intersect(column_aliases[[std]], names(lookup))
		if (length(hit) > 0) {
			out[[std]] <- df[[lookup[[hit[1]]]]]
		}
	}
	out <- as_tibble(out)
	missing <- setdiff(c("country", "score"), names(out))
	if (length(missing) > 0) {
		stop(path, ": could not find column(s) ", paste(missing, collapse = ", "))
	}
	if (!"year" %in% names(out)) {
		year <- str_extract(basename(path), "(19|20)\\d{2}")
		if (is.na(year)) {
			stop(path, ": no 'year' column and no year in the file name")
		}
		out$year <- as.integer(year)
	}
	out
}
```

## Data

```{r}
#| label: load
world <- map(data_files, load_file) |>
	bind_rows() |>
	mutate(
		year = as.integer(year),
		country = str_trim(as.character(country)),
		canonical = unname(alias_to_country[tolower(country)]),
		is_sea = !is.na(canonical),
		country = coalesce(canonical, country)
	) |>
	select(-canonical)

sea <- filter(world, is_sea)
if (nrow(sea) == 0) stop("No Southeast Asian countries found in the data.")

factors <- intersect(all_factors, names(sea))
latest_year <- max(sea$year)
multi_year <- n_distinct(sea$year) > 1
missing <- setdiff(names(sea_countries), unique(sea$country))
```

The data has **`r nrow(sea)` Southeast Asian rows** covering
**`r n_distinct(sea$country)` countries** from `r min(sea$year)` to
`r latest_year`.
`r if (length(missing) > 0) paste0("Not in the data: ", paste(missing, collapse = ", "), ".")`

## Ranking in `r latest_year`

```{r}
#| label: ranking-table
world_latest <- world |>
	filter(year == latest_year) |>
	mutate(world_rank = min_rank(desc(score)))

ranking <- world_latest |>
	filter(is_sea) |>
	arrange(desc(score)) |>
	mutate(sea_rank = row_number()) |>
	select(sea_rank, world_rank, country, score, all_of(factors))

no_latest <- setdiff(unique(sea$country), ranking$country)

# Factor values per country are in the heatmap below.
ranking |>
	select(sea_rank, world_rank, country, score) |>
	kable(
		col.names = c("Regional rank", "World rank", "Country", "Happiness score"),
		digits = 2
	)
```

`r if (length(no_latest) > 0) paste0("No ", latest_year, " data for ", paste(no_latest, collapse = ", "), "; they are left out of the ranking, heatmap and scatter plot.")`

```{r}
#| label: fig-ranking
#| fig-cap: "Happiness score by country in the latest year."
ggplot(ranking, aes(x = score, y = reorder(country, score))) +
	geom_col(fill = blue, width = 0.6) +
	geom_text(aes(label = sprintf("%.2f", score)), hjust = -0.2,
		size = 3.2, colour = text_muted) +
	scale_x_continuous(limits = c(0, max(10, max(ranking$score) + 0.8)),
		expand = expansion(mult = c(0, 0.02))) +
	labs(x = "Happiness score (0-10)", y = NULL) +
	theme(panel.grid.major.y = element_blank())
```

## Southeast Asia vs. the rest of the world

Mean of each measure in `r latest_year`.

```{r}
#| label: sea-vs-world
world_latest |>
	select(is_sea, score, all_of(factors)) |>
	pivot_longer(-is_sea, names_to = "measure") |>
	group_by(measure, is_sea) |>
	summarise(mean = mean(value, na.rm = TRUE), .groups = "drop") |>
	pivot_wider(names_from = is_sea, values_from = mean) |>
	transmute(
		Measure = factor_labels[measure],
		`Southeast Asia` = `TRUE`,
		`Rest of world` = `FALSE`,
		Difference = `TRUE` - `FALSE`,
		order = match(measure, names(factor_labels))
	) |>
	arrange(order) |>
	select(-order) |>
	kable(digits = 3)
```

## Happiness factors by country

```{r}
#| label: fig-heatmap
#| fig-cap: "Factor values per country (rows ordered by happiness score). Shading is relative to the region: lightest = lowest, darkest = highest."
#| fig-width: 9
#| fig-height: 5.5
ranking |>
	select(country, score, all_of(factors)) |>
	pivot_longer(all_of(factors), names_to = "factor") |>
	group_by(factor) |>
	mutate(scaled = (value - min(value, na.rm = TRUE)) /
		(max(value, na.rm = TRUE) - min(value, na.rm = TRUE))) |>
	ungroup() |>
	mutate(
		factor = factor(factor_labels[factor], levels = factor_labels[factors]),
		country = reorder(country, score)
	) |>
	ggplot(aes(x = factor, y = country, fill = scaled)) +
	geom_tile(colour = "white", linewidth = 1) +
	geom_text(aes(label = sprintf("%.2f", value),
		colour = scaled > 0.6), size = 3) +
	scale_fill_distiller(palette = "Blues", direction = 1, limits = c(0, 1),
		name = "Relative to\nregion") +
	scale_colour_manual(values = c(`TRUE` = "white", `FALSE` = "#0b0b0b"),
		guide = "none") +
	labs(x = NULL, y = NULL) +
	theme(
		panel.grid = element_blank(),
		axis.text.x = element_text(angle = 30, hjust = 1)
	)
```

## What goes with happiness?

Correlation of each factor with the happiness score, across all Southeast Asian
rows in the data.

```{r}
#| label: fig-correlations
#| fig-cap: "Correlation of each factor with the happiness score. Blue = positive, orange = negative."
#| fig-width: 7
#| fig-height: 4
corr <- tibble(
	factor = factors,
	r = map_dbl(factors, \(f) cor(sea[[f]], sea$score, use = "complete.obs"))
) |>
	mutate(label = factor_labels[factor])

ggplot(corr, aes(x = r, y = reorder(label, r), fill = r >= 0)) +
	geom_col(width = 0.6) +
	geom_vline(xintercept = 0, colour = gray) +
	geom_text(aes(label = sprintf("%+.2f", r), hjust = ifelse(r >= 0, -0.2, 1.2)),
		size = 3.2, colour = text_muted) +
	scale_fill_manual(values = c(`TRUE` = blue, `FALSE` = orange), guide = "none") +
	scale_x_continuous(limits = c(-1.15, 1.15)) +
	labs(x = "Correlation with happiness score", y = NULL) +
	theme(panel.grid.major.y = element_blank())
```

```{r}
#| label: fig-gdp
#| fig-cap: "GDP per capita vs. happiness score in the latest year."
#| fig-width: 7.5
if ("gdp" %in% factors) {
	ranking |>
		filter(!is.na(gdp), !is.na(score)) |>
		ggplot(aes(x = gdp, y = score)) +
		geom_point(colour = "white", size = 4) +
		geom_point(colour = blue, size = 3) +
		geom_text(aes(label = country), hjust = -0.15, vjust = -0.4,
			size = 3.2, colour = text_muted) +
		scale_x_continuous(expand = expansion(mult = c(0.05, 0.15))) +
		labs(x = "GDP per capita (as reported in the source data)",
			y = "Happiness score")
}
```

## Trends over time

Happiness score per country (blue) against the Southeast Asian mean (gray).

```{r}
#| label: trends
#| results: asis
#| fig-width: 10
#| fig-height: 7
if (!multi_year) {
	cat("The data covers a single year, so there are no trends to show.")
} else {
	regional <- sea |>
		group_by(year) |>
		summarise(score = mean(score, na.rm = TRUE))

	ggplot(sea, aes(x = year, y = score)) +
		geom_line(data = regional, colour = gray, linewidth = 0.9) +
		geom_line(colour = blue, linewidth = 0.9) +
		geom_point(colour = blue, size = 1) +
		facet_wrap(~ country, ncol = 4) +
		scale_x_continuous(breaks = scales::breaks_pretty(4)) +
		labs(x = NULL, y = "Happiness score") +
		theme(strip.text = element_text(hjust = 0, size = 10))
}
```

```{r}
#| label: score-change
if (multi_year) {
	sea |>
		group_by(country) |>
		arrange(year, .by_group = TRUE) |>
		summarise(
			`First year` = first(year), `First score` = first(score),
			`Last year` = last(year), `Last score` = last(score)
		) |>
		mutate(Change = `Last score` - `First score`) |>
		arrange(desc(Change)) |>
		rename(Country = country) |>
		kable(digits = 2)
}
```
